---
title: "PiGx: BS-seq"
output: 
  html_notebook:
    toc:        TRUE
    toc_float:  TRUE
    theme:      "lumen"
    number_sections: FALSE
    code_folding: "hide"
    self_contained: TRUE
    includes:
      in_header: pigx_bsseq_logo.html
output_dir: "/Users/agosdsc/Development/Snakemake/pigx_bsseq/annotation_rule/"
output_file: "sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted.annotation.nb.html"
date: "`r Sys.Date()`"
#bibliography: reports.bib
params:
  inBed:    "/home/kwreczy/tmp/my_output/differential_methylation/0_1.sorted_diffmeth.bed"
  assembly: "ce10"
  refseqfile: "/blabla/annotation/refseq.genes.ce10.bed"
  grFile: "ablabla.RData"
  grFile_hyper:    "/blala.RData"
  grFile_hypo:    "/blabla.RData"
  ideoDMC_script: "/blabla/scripts/ideoDMC.R"
---

```{r eval_params, echo=FALSE}

inBed      <- params$inBed
assembly      <- params$assembly
refgenes.file <- params$refseqfile
grFile      <- params$grFile
grFile_hyper      <- params$grFile_hyper
grFile_hypo      <- params$grFile_hypo
ideoDMC_script <- params$ideoDMC_script

```

## Annotation of differentially methylated bases


```{r, include=FALSE,warning=FALSE}

  suppressPackageStartupMessages(expr = {
    library("genomation")
    library("rtracklayer")
    library("DT")
    library("ggplot2")
    library("GenomeInfoDb")

  })
```

### Libraries
```{r, eval=FALSE, include=TRUE}

    library("genomation")
    library("rtracklayer")
    library("DT")
    library("ggplot2")
    library("GenomeInfoDb")

```

### Parameters
```{r print_params, echo=TRUE, results="asis"}

paramsDesc <- c("Input Data (diff.methyl.)",
                "Input RDS Data (diff.methyl.)",
                "Input RDS Data (hyper-methyl.)",
                "Input RDS Data (hypo-methyl.)",
                "Assembly",
                "RefSeq genes file")

paramsValues <- c( inBed ,
                   grFile,
                   grFile_hyper,
                   grFile_hypo,
                   assembly,
                   refgenes.file )

inputParams <- data.frame(Parameters = paramsDesc,
                          Values = paramsValues,
                          stringsAsFactors = FALSE)

knitr::kable(inputParams)

```


### Preprocessing

```{r, message=FALSE, echo=TRUE}


# Load differentially methylated bases from a BED file.
GRanges.obj <- readBed(file = inBed,
                      track.line = "auto")
# Load differentially methylated, hypo- and hyper-methylated
# bases from RDS files.
methylDiff.obj <- readRDS(grFile)
methylDiff.obj_hyper = readRDS(grFile_hyper)
methylDiff.obj_hypo = readRDS(grFile_hypo)

if( !is.null(methylDiff.obj_hypo)){
  GRanges.obj_hypo = as(methylDiff.obj_hypo, "GRanges")
}else{
  GRanges.obj_hypo = GRanges()
}

if( !is.null(methylDiff.obj_hyper)){
  GRanges.obj_hyper = as(methylDiff.obj_hyper, "GRanges")
}else{
  GRanges.obj_hyper = GRanges()
}


```


We query the UCSC table browser directly to fetch the reference gene set and CpG islads locations for the given genome assembly using the [rtracklayer](http://bioconductor.org/packages/release/bioc/html/rtracklayer.html) Package [@rtracklayer2009]. After we have aquired the reference genes, we can determine different features intrinsic to each gene, such as promoter, exon or intron using the `readTranscriptFeatures()` function from [genomation](http://bioinformatics.mdc-berlin.de/genomation/) [@genomation2014] and appropriately
for CpG islands and shores.


```{r annotate with refseq genes}

#' Fetch a table from the UCSC browser and save it as a BED file
#' 
#' The windows are color coded based on their score (methylation or differential
#' methylation value).
#' 
#' @param table.name a character indicating name of a UCSC table
#' @param table.loc a character indicating path to the output BED file
#' @param assembly a character indicating a genome version, e.g. "ce10""
#' 
#' @return location of a output BED file which can be visualized in the UCSC browser 
#' 
#' 
#' @export
#' @docType methods
#' @rdname fetchTablefromUCSC
fetchTablefromUCSC <- function(table.name,
                       table.loc = NULL,
                       assembly){
  
      if(is.null(table.loc)) table.loc <- paste0(table.name, assembly,".bed")
  
      ## import local bed file if available
      if( file.exists(table.loc) ) {
          ## parse it 
          message(paste0("Found ",table.name," track at:\n", table.loc))
          return(table.loc)
    
      } else {
        message("Trying to fetch from UCSC table browser.\n")
        ## so we check with rtracklayer for the latest ucsc data
        require(rtracklayer)
        mySession = browserSession("UCSC")
        genome(mySession) <- assembly
        track.names <- trackNames(ucscTableQuery(mySession))
        
        if(table.name %in% track.names) {
          message(paste0("Found single ",table.name," track, downloading...\n"))
          # fetch it as a GRanges object 
          targetTrack <- track(mySession,table.name)
          ## and write it to BED file
          export.bed(object = targetTrack,
                     con = table.loc,
                     trackLine=FALSE)
          
          message(paste0("Written the RefSeq track to:\n",table.loc))
          return(table.loc)
        } else {
          print(paste("Could not find ",table.name," for the given assembly <'",assembly,"'>." ))
          return(NA)
        }
      }
}

## filename can either be passed as argument or be this generic pasted one
refgenes.file <- paste0("refseq.genes.",assembly,".bed")
fetched.refgenes = fetchTablefromUCSC("refGene", table.loc = refgenes.file, assembly)

cpgi.file <- paste0("cpgi.",assembly,".bed")
fetched.cpgi = fetchTablefromUCSC("cpgIslandExt", table.loc = cpgi.file, assembly)

```


### Show number of differentially methylated bases per chromosome

```{r, message=FALSE, echo=TRUE}

number.of.elements.per.chr = function(gr.obj){
  
  # Optional, if you want a genomic order of the chromosomes
  diffmeth.gr = sortSeqlevels(gr.obj)
  # Split into a GRangesList
  # where each element has all ranges for one chromosome
  diffmeth.gr.perchr = split(diffmeth.gr, seqnames(diffmeth.gr))
  el.chrs = elementNROWS(diffmeth.gr.perchr)
  
  df = data.frame(Chromosome=names(el.chrs), Number.diff.meth.bases=as.vector(el.chrs))
  return(df)
}

# Show number of differentially methylated bases per chromosome
if(nrow(GRanges.obj)!=0){
   stats.df = number.of.elements.per.chr(GRanges.obj)
   datatable(stats.df,
          extensions = 'Buttons',
          options = list(
            dom = 'Brtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Number of differentially methylated bases per chromosome",
          rownames = FALSE)
}else{
  print("There are no differentially methylated bases.")
}


if(length(GRanges.obj_hyper)!=0){
   stats.df = number.of.elements.per.chr(GRanges.obj_hyper)
   datatable(stats.df,
          extensions = 'Buttons',
          options = list(
            dom = 'Brtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Number of hyper-methylated bases per chromosome",
          rownames = FALSE)
}else{
  print("There are no hyper-methylated bases.")
}


if(length(GRanges.obj_hypo)!=0){
   stats.df = number.of.elements.per.chr(GRanges.obj_hypo)
   datatable(stats.df,
             extensions = 'Buttons',
          options = list(
            dom = 'Brtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Number of hypo-methylated bases per chromosome",
          rownames = FALSE)
}else{
  print("There are no hypo-methylated bases.")
}

```


### Annotation

Now we can load differentially methylated bases to detect overlap percentage with gene features, CpG islands and shores.

```{r, message=FALSE,fig.show="hold"}

if(!is.na(fetched.refgenes) & nrow(GRanges.obj)!=0){
  ## now we parse the gene features
  refgenes.grl <- readTranscriptFeatures(refgenes.file)

  annot.gene <- annotateWithGeneParts(target = GRanges.obj,
                                         feature = refgenes.grl,
                                         intersect.chr = TRUE)

  # plot the target overlap for each
  genomation::plotTargetAnnotation(annot.gene,
                       main="Differential methylation annotation")
}

if(!is.na(fetched.cpgi) & nrow(GRanges.obj)!=0){
  
  # read the shores and flanking regions and name the flanks as shores 
  # and CpG islands as CpGi
  cpg.obj=readFeatureFlank(cpgi.file,
                           feature.flank.name=c("CpGi","shores"))
  #
  # convert methylDiff object to GRanges and annotate
  diffCpGann=annotateWithFeatureFlank(GRanges.obj,
                                      cpg.obj$CpGi, cpg.obj$shores,
                                      feature.name="CpGi",flank.name="shores")

  # plot the target overlap for each
  genomation::plotTargetAnnotation(diffCpGann, 
                       col=c("green","gray","white"),
                       main="Differential methylation annotation")
}


```


### Plot distance to the nearest TSSs

After getting the annotation of differentially methylated regions, we can get the distance to TSS and the nearest gene name using the  `getAssociationWithTSS` function from genomation package.

```{r, message=FALSE, echo=TRUE}

if(!is.na(fetched.refgenes) & nrow(GRanges.obj)!=0){

# Get distance to nearest TSS and gene id from AnnotationByGeneParts
# target.row is the row number in diffmeth.gr
assoTSS = getAssociationWithTSS(annot.gene)

datatable(as.data.frame(assoTSS),
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Distance to nearest TSS",
          rownames = FALSE)


hist(assoTSS[,2], main="Distance to the nearest TSS", xlab="bp", ylab="Frequency", col="cornflowerblue")

}


```



### Plot overview of hyper and hypo-methylated regions over the genome

```{r, message=FALSE,fig.show="hold"}

if(length(GRanges.obj_hypo)>1 & length(GRanges.obj_hyper)>1){

chr.len.df=fetchExtendedChromInfoFromUCSC(assembly,
        goldenPath_url="http://hgdownload.cse.ucsc.edu/goldenPath",
        quiet=FALSE)
chr.len=chr.len.df[,2]
names(chr.len)=chr.len.df[,1]


source(ideoDMC_script)
plot( ideoDMC_hyper_hypo(methylDiff.obj_hyper, methylDiff.obj_hypo, chrom.length = chr.len,
   circos = TRUE, title = "Differential methylation", hyper.col = "magenta", hypo.col = "green") 
   )

plot( ideoDMC_hyper_hypo(methylDiff.obj_hyper, methylDiff.obj_hypo, chrom.length = chr.len,
   circos = FALSE, title = "Differential methylation", hyper.col = "magenta", hypo.col = "green") )

}

```



## Session Info

```{r }
sessionInfo()
```



## References