---
title: "PiGx: BS-seq"
output: 
  html_notebook:
    toc:        TRUE
    toc_float:  TRUE
    theme:      "lumen"
    number_sections: FALSE
    code_folding: "hide"
    self_contained: TRUE
    includes:
      in_header: pigx_bsseq_logo.html
date: "`r Sys.Date()`"
bibliography: reports.bib
params:
  inputfiles: ["methylation_calls/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_methylRaw.RData","methylation_calls/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_methylRaw.RData"]
  grdata: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RData"
  outBed:   "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows.bed"
  sampleids: ["sampleA", "sampleB"]
  assembly: "ce10"
  treatment: [0,1]
  mincov: 0
  context: "CpG"
  workdir: "~/outputdir/"
  cores: 2
---
  
  
```{r eval_params, echo=FALSE}

input     <- params$inputfiles
sampleids <- params$sampleids
output    <- params$outBed
grFile    <- params$grdata
assembly  <- params$assembly
treatment <- params$treatment
mincov    <- params$mincov
context   <- params$context
workdir   <- params$workdir
cores     <- params$cores

```

## Differential methylation report

  TODO: add few sentences about diff meth.

```{r, include=FALSE,warning=FALSE}

suppressPackageStartupMessages(expr = {
  ## load methylKit
  library("methylKit",warn.conflicts = FALSE,quietly = TRUE)
  ## load DT 
  library("DT")
})
```

### Libraries
```{r load libraries, eval=FALSE, include=TRUE}

## load methylKit
library("methylKit")

```

### Parameters
```{r print_params, echo=TRUE}


print("PARAMS")
print(params)

# convert treatment values to integers


# paramsDesc <- c("Input Data",
#                 "Output file")
# 
# paramsValues <- c( input,
#                    output)
# 
# print("paramsValues")
# print(paramsValues)
# 
# inputParams <- data.frame(Parameters = paramsDesc,
#                           Values = paramsValues,
#                           stringsAsFactors = FALSE)

#inputList <- list(InputData=input, Outputfile=output)

#params
#datatable(inputParams,options = list(dom = 't'),rownames = FALSE)
#knitr::kable(inputList)
```

### Differential methylation

#TODO

```{r differential methylation, echo=TRUE}

inputfiles = paste0(workdir, input)

###################################### TODO: this is ugly, improvee it
# becauses not treatment in the input file can be only numbers...
treatment = as.numeric(treatment)
###################################### 

methRawList.obj = methRead(as.list(inputfiles),
                                sample.id = as.list(sampleids),
                                assembly = assembly,
                                treatment=treatment,
                                context = context,
                                mincov = mincov
)
print(methRawList.obj)

# Take bases with coverage from all samples are retained
meth.unite=unite(methRawList.obj, destrand=FALSE)



################ TODO:
# onyl CpG context now
#covariates
# how many cores for diff meth
cores=2
# which test to use
################


if(nrow(meth.unite)!=0){
  
  meth.diffmeth <- calculateDiffMeth(meth.unite, 
                                   overdispersion="MN",
                                   adjust = "qvalue",
                                   test="Chisq",
                                   mc.cores=cores)
}


```


### Export

Finally we export the regions to a *BED* file, which can be loaded into any genome browser like [IGV](http://software.broadinstitute.org/software/igv/) or [UCSC](https://genome.ucsc.edu/) to allow for further analysis, annotation and visualisation. 
```{r export to bed}

#' Export windows to a BED file
#' 
#' The windows are color coded based on their score (methylation or differential
#' methylation value).
#' 
#' @param windows \code{\link[GenomicRanges]{GRanges}} object with information about
#' differentially methylated regions
#' @param filename name of the output data
#' @param trackLine UCSC browser trackline
#' @param colramp color scale to be used in the BED display
#' defaults to gray,green, darkgreen scale.
#' 
#' @return A BED files with the differentially methylated regions
#' which can be visualized in the UCSC browser 
#' 
#' @seealso \code{\link{methylKit::methSeg2bed}}
#' 
#' @export
#' @docType methods
#' @rdname meth2bed
meth2bed<-function(windows,filename,
    trackLine="track name='meth windows' description='meth windows' itemRgb=On",
    colramp=colorRamp(c("gray","green", "darkgreen"))
                        ){
  if(class(windows)!="GRanges"){
    stop("windows object has to be of class GRanges")
  }
  
  ## case if only one line is exported
  if(is.null(colramp) | length(windows)==1){
    trackLine <- gsub(pattern = "itemRgb=On",replacement = "",x = trackLine)
  } else {
    require(rtracklayer)
    ramp <- colramp
    scores=windows$meth.diff
    mcols(windows)$itemRgb= rgb(ramp(scores), maxColorValue = 255)     
  }
  
  strand(windows)="*"
  score(windows)=windows$meth.diff
  
  if(is.null(trackLine)){
    
    export.bed(windows,filename)
  }else{
    export.bed(windows,filename,
               trackLine=as(trackLine, "BasicTrackLine"))
  }
}


if(nrow(meth.unite)!=0){
  
  # Convert a methylDiff object to a GRanges object
  meth.diffmeth.gr = as(meth.diffmeth, "GRanges")

  # Export differentially methylated regions to a bed file
  meth2bed(windows = meth.diffmeth.gr,
            trackLine = paste0("track name='differentially methylated regions ' ",
                               "description='diff. meth. between ",
                               paste(meth.diffmeth@sample.ids,collapse=","),
                               " mapped to ",
                               meth.diffmeth@assembly,
                               "' itemRgb=On"),
            colramp=colorRamp(c("gray","green", "darkgreen")),
            filename = output)

}

```

The GRanges object is saved to enable fast loading for further analysis.
```{r}
if(nrow(meth.unite)!=0){

## Saving object
saveRDS(meth.diffmeth.gr, grFile)
  
}
```


```{r, echo=FALSE}

#a=meth.unite[-(1:674),]
# If samples
#touch file to force snakemake to exit in nice way and generate html anyways ??? #TODO
# create a 
if(nrow(meth.unite)==0){
  
  trackLine = paste0("track name='differentially methylated regions ' ",
                               "description='diff. meth. between ",
                               paste(meth.unite@sample.ids,collapse=","),
                               " mapped to ",
                               meth.unite@assembly,
                               "' itemRgb=On")
  fileConn<-file(output)
  writeLines(trackLine, fileConn)
  close(fileConn)
  
  saveRDS(as(meth.unite, "GRanges"), grFile)
}

```



## Session Info

```{r sessioninfo}
sessionInfo()
```



## References