---
title: "PiGx: BS-seq"
output: 
  html_notebook:
    toc:        TRUE
    toc_float:  TRUE
    theme:      "lumen"
    number_sections: FALSE
    code_folding: "hide"
    self_contained: TRUE
    includes:
      in_header: pigx_bsseq_logo.html
date: "`r Sys.Date()`"
bibliography: reports.bib
params:
  inputfiles: ["methylation_calls/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_methylRaw.RData","methylation_calls/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_methylRaw.RData"]
  grdata: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RData"
  grdata_hyper: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RData"
  grdata_hypo: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RData"
  outBed:   "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows.bed"
  sampleids: ["sampleA", "sampleB"]
  assembly: "ce10"
  treatment: [0,1]
  mincov: 0
  context: "CpG"
  workdir: "~/outputdir/"
  cores: 2
---
  
  
```{r eval_params, echo=FALSE}

input     <- params$inputfiles
sampleids <- params$sampleids
output    <- params$outBed
grFile    <- params$grdata
grFile_hyper    <- params$grdata_hyper
grFile_hypo    <- params$grdata_hypo
assembly  <- params$assembly
treatment <- params$treatment
mincov    <- params$mincov
context   <- params$context
workdir   <- params$workdir
cores     <- params$cores

```

# Differential methylation report

## Calling differentially methylated bases

The logistic regression is applied to model the log odds ratio which is based on methylation proportion of a CpG using the treatment vector which denotes the sample group membership for the CpGs in the model. 

When covariates (such as e.g. age) are included in the analysist, we will test if full model (model with treatment and covariates) is better than the model with the covariates only. If there is no effect due to the treatment (sample groups), the full model will not explain the data better than the model with covariates only.

After q-value calculation, differentially methylated bases are extracted based on q-value and percent methylation difference cutoffs. Here we select bases that have q-value < 0.01 and percent methylation difference larger than 25%. Futhermore, we calculate hyper-methylated or hypo-methylated bases. 

Overdispersion occurs when there is more variability in the data than assumed by the distribution and is here included in the differentially methylatation calculation.

For more details about `calculateDiffMeth()` and `getMethylDiff()` functions see [@methylKit2012] and for details about
the logistic regression and overdispersion see [@Wreczycka2017].


```{r, include=FALSE,warning=FALSE}

suppressPackageStartupMessages(expr = {
  ## load methylKit
  library("methylKit",warn.conflicts = FALSE,quietly = TRUE)
  library("DT",warn.conflicts = FALSE,quietly = TRUE)
})

```

### Libraries
```{r load libraries, eval=FALSE, include=TRUE}

## load libraries
library("methylKit")
library("DT")

```

### Parameters
```{r print_params, echo=TRUE}

inputParams <- data.frame(sampleids=sampleids,
                          input1 = input[1],
                          input2 = input[2],
                          treatment=treatment,
                          context=context,
                          mincov=mincov,
                          assembly=assembly,
                          output=output,
                          RDSfile=grFile,
                          stringsAsFactors = FALSE)

datatable(inputParams,
          options = list(dom="t", autoWidth = TRUE, scrollX=TRUE, colReorder = TRUE),
          rownames = FALSE, class = 'cell-border stripe')

```

### Load methylation profiles


```{r load methylation, echo=TRUE,warning=FALSE}

###################################### TODO: this is ugly, improvee it
# becauses not treatment in the input file can be only numbers...
treatment = as.numeric(treatment)
################ TODO:
cores=2
################
###################################### 

inputfiles = paste0(workdir, input)
methRawList.obj = methRead(as.list(inputfiles),
                                sample.id = as.list(sampleids),
                                assembly = assembly,
                                treatment=treatment,
                                context = context,
                                mincov = mincov
)

# Take bases with coverage from all samples are retained
meth.unite=unite(methRawList.obj, destrand=FALSE)

############### onyl for now, remove it later ################################################
# saveRDS(methRawList.obj, "~/tmp/methRawList.obj.rds")
# 
# #methRawList.obj = readRDS("~/tmp/methRawList.obj.rds")
# 
# data(methylKit)
# 
# covariates=data.frame(age=c(30,80,30,80))
# meth.unite<-dataSim(replicates=4,sites=1000,treatment=c(1,1,0,0),
#                              covariates=covariates,
#                              sample.ids=c("test1","test2","ctrl1","ctrl2"))

###############  ################################################


```

### Find differentially methylated bases


```{r, differential methylation, echo=TRUE,warning=FALSE}

# Cutoff for absolute value of methylation percentage change
# between test and control
difference = 25
# Cutoff for qvalue of differential methylation statistic
qvalue = 0.01

if(nrow(meth.unite)!=0){
  
  meth.diffmeth <- calculateDiffMeth(meth.unite, 
                                   overdispersion="MN",
                                   adjust = "qvalue",
                                   test="Chisq",
                                   mc.cores=cores)

  # Get differentially methylated regions/bases based on cutoffs  
  meth.diffmeth.all = getMethylDiff(meth.diffmeth, 
                                           difference=difference,
                                           type="all",
                                           qvalue=qvalue)
  # Get hypo-methylated
  meth.diffmeth.hypo = getMethylDiff(meth.diffmeth,
                                           difference=difference,
                                           type="hypo",
                                           qvalue=qvalue)
  # Get hyper-methylated
  meth.diffmeth.hyper = getMethylDiff(meth.diffmeth,
                                           difference=difference,
                                           type="hyper",
                                           qvalue=qvalue)
}


methylDiff2df_roundint = function(methylDiff.obj, roundval=4){
  df = as(methylDiff.obj, "data.frame")
  df[,5:7] = round(df[,5:7], roundval)
  return(df)
}

## show the regions
datatable(methylDiff2df_roundint(meth.diffmeth),
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "All regions",
          rownames = FALSE)


## show differentially methylated regions
if(nrow(meth.diffmeth.all)!=0){
datatable(methylDiff2df_roundint(meth.diffmeth.all),
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Differentially methylated bases",
          rownames = FALSE)
}

## show hyper-methylated region
if(nrow(meth.diffmeth.hyper)!=0){
datatable(methylDiff2df_roundint(meth.diffmeth.hyper),
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Hyper-methylated bases",
          rownames = FALSE)
}

## show hypo-methylated region
if(nrow(meth.diffmeth.hypo)!=0){
datatable(methylDiff2df_roundint(meth.diffmeth.hypo),
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE),
          caption = "Hypo-methylated bases",
          rownames = FALSE)
}
  
```


### Export differentially methylated bases

Finally we export the regions to a *BED* file, which can be loaded into any genome browser like [IGV](http://software.broadinstitute.org/software/igv/) or [UCSC](https://genome.ucsc.edu/) to allow for further analysis, annotation and visualisation. 
```{r export to bed,warning=FALSE}

#' Export windows to a BED file
#' 
#' The windows are color coded based on their score (methylation or differential
#' methylation value).
#' 
#' @param windows \code{\link[GenomicRanges]{GRanges}} object with information about
#' differentially methylated regions
#' @param filename name of the output data
#' @param trackLine UCSC browser trackline
#' @param colramp color scale to be used in the BED display
#' defaults to gray,green, darkgreen scale.
#' 
#' @return A BED files with the differentially methylated regions
#' which can be visualized in the UCSC browser 
#' 
#' @seealso \code{\link{methylKit::methSeg2bed}}
#' 
#' @export
#' @docType methods
#' @rdname meth2bed
meth2bed<-function(windows,filename,
    trackLine="track name='meth windows' description='meth windows' itemRgb=On",
    colramp=colorRamp(c("gray","green", "darkgreen"))
                        ){
  
  range01 <- function(x){(x-min(x)+1)/(max(x)-min(x)+1)}

  if(class(windows)!="GRanges"){
    stop("windows object has to be of class GRanges")
  }
  
  ## case if only one line is exported
  if(is.null(colramp) | length(windows)==1){
    trackLine <- gsub(pattern = "itemRgb=On",replacement = "",x = trackLine)
  } else {
    require(rtracklayer)
    ramp <- colramp
    score(windows)=range01(windows$meth.diff)
    mcols(windows)$itemRgb= rgb(ramp(score(windows)), maxColorValue = 255)     
  }
  
  strand(windows)="*"
  score(windows)=range01(windows$meth.diff)
  
  if(is.null(trackLine)){
    
    export.bed(windows,filename)
  }else{
    export.bed(windows,filename,
               trackLine=as(trackLine, "BasicTrackLine"))
  }
}



# Export differentially methylated bases

if(nrow(meth.diffmeth)!=0){
  

####################################################### TODO: change it to meth.diffmeth.all!!!!
  
  # Convert a methylDiff object to a GRanges object
  meth.diffmeth.gr = as(meth.diffmeth, "GRanges")
  
  # Export differentially methylated bases to a bed file
  trackLine = paste0("track name='differentially methylated bases ' ",
                               "description='diff. meth. between ",
                               paste(meth.diffmeth@sample.ids,collapse=","),
                               " mapped to ",
                               meth.diffmeth@assembly,
                               "' itemRgb=On")
  meth2bed(windows = meth.diffmeth.gr,
            trackLine=trackLine,
            colramp=colorRamp(c("gray","green", "darkgreen")),
            filename = output)

}


```

The GRanges object is saved to enable fast loading for further analysis.
```{r}


if(nrow(meth.unite)!=0){

####################################################### TODO: change it to meth.diffmeth.all!!!!

  ## Saving object from methylDiff to a RDS file
  saveRDS(meth.diffmeth, grFile)
  saveRDS(meth.diffmeth.hyper, grFile_hyper)
  saveRDS(meth.diffmeth.hypo, grFile_hypo)
  
}
```


```{r, echo=FALSE}

# In case when there are no differentially methylated bases
# create an empty output

if(nrow(meth.unite)==0){
  
  trackLine = paste0("track name='differentially methylated bases ' ",
                               "description='diff. meth. between ",
                               paste(meth.unite@sample.ids,collapse=","),
                               " mapped to ",
                               meth.unite@assembly,
                               "' itemRgb=On")
  fileConn<-file(output)
  writeLines(trackLine, fileConn)
  close(fileConn)
  
  saveRDS(as(meth.unite, "GRanges"), grFile)
  saveRDS(as(meth.unite, "GRanges"), grFile_hyper)
  saveRDS(as(meth.unite, "GRanges"), grFile_hypo)
}

```


## Session Info

```{r sessioninfo}
sessionInfo()
```



## References